---
title: "Predicting_Students_Dropout_and_Academic_Success"
author: "James Kibiro, Josh Balderas, Neelkumar Patel"
format: html
editor: visual
---

## Predicting Students Dropout and Academic Success

Research Question: Does age at enrollment contribute to student academic dropout and failure in higher education?

```{r}
# set up the libraries
library(tidyverse)
library(readxl) 
library(here) # sets path to your folder directory
library(leaps) # for variable selection methods
library(ggplot2)
library(caret)
library(heatmaply)
```

```{r}
# load the data
pj_data <- read_xlsx(here("data", "project_data.xlsx")) # the data file is in a folder named 'data'
head(pj_data)
str(pj_data)
```

```{r}
summary(pj_data)
pj_data <- na.omit(pj_data) # remove any missing values
```

```{r}
# Factorize the categorical variables
# 'Target' is the only categorical variable
col <- c("Target")
pj_data [, col] <- lapply(pj_data[, col], factor)
str(pj_data) # confirm categorization of the column
```

For our research question, "Does age at enrollment contribute to student academic dropout and failure in higher education?", we will take 'Age at Enrollment' as the dependent (Y) variable that we can predict using the other variables as the independent variables.

```{r}
# Let's take a look at the distribution of our variable of interest
pj_data |>
  select(`Age at enrollment`)|>
  ggplot(aes(x = `Age at enrollment`))+
           geom_histogram(bins = 50, fill = "steelblue", color = "white")+
  theme_minimal()+
  labs( title = "Histogram of Age at Enrollment")+
  geom_rug() # show the full breadth of the distribution
```

Let's develop a multivariate linear regression model to predict student academic performance.

Since we have many independent variables, a multivariate linear regression would be ideal. We have 37 variables and therefore, we can 36 (i.e.,37-1) independent variables.

Since the outcome (Y) is 'Age at enrollment' what we want to predict Let's split the data 70% training and 30% testing

```{r}
set.seed(123)  # setting a seed for reproducibility of the random split
index <- createDataPartition(pj_data$`Age at enrollment`, p = 0.3, list = FALSE)
training_data <- pj_data[-index, ]
testing_data <- pj_data[index, ]
```

Check summary statistics of the outcome "Age at enrollment" in training and testing data

```{r}
summary(training_data$`Age at enrollment`)
summary(training_data$`Age at enrollment`)
```

Let's develop an MLR to predict 'Age at enrollment' using all other 36 variables as independent variables

```{r}
model_0 <- lm(`Age at enrollment`~., data = training_data)
summary(model_0)
```

Adjusted R-squared: 0.5775 low and 20 of the independent variables are significant at the 0.05 significance level

checking model assumptions

-   Normality: Residuals are normally distributed with a mean of 0

-   Linear relationship between X and Y --\> check residual plot

-   Constant variance among residuals: Linear regression assumes the variance among error terms are constant (also called: homoscedasticity).

check normality assumption by plotting the histogram of residuals and how much it resembles a normal distribution/

```{r}
##### 
res0_df <- data.frame(res = as.numeric(model_0$residuals))

# check normality assumption by plotting the histogram of residuals

ggplot(data = res0_df, aes( x = res))+
  geom_histogram(bins = 50, fill = "steelblue", color = "white")+
  theme_minimal()+
  labs( title = "Histogram of Residuals")+
  geom_rug(color = "steelblue") # show the full breadth of the distribution

```

We see a normal distribution.

Let's check for normality assumption using Quantile-Quantile (QQ) plots: It’s good if residuals follow the straight dashed line.

```{r}
res0_df |>
  ggplot(aes(sample = res))+
  geom_qq(color = "steelblue")+
  geom_qq_line(color = "orange")+
  labs( title = "QQ Plot of Residuals vs the theoretical normal distribution")+
  theme_minimal()

```

We see a huge variation from the diagonal line at the extremes, we could try data transformation (e.g., log transform the outcome data)

```{r}
model_1 <- lm(log(`Age at enrollment`)~., data = training_data)
res1_df <- data.frame(res1 = as.numeric(model_1$residuals))

res1_df |>
  ggplot(aes(sample = res1))+
  geom_qq(color = "steelblue")+
  geom_qq_line(color = "orange")+
  labs( title = "QQ Plot of Residuals after log transformation")+
  theme_minimal()

```

The variation is still evident with the data transformation; therefore normality assumption is violated.

Residual Plot: Residuals vs Fitted visual -- check the linear relationship assumptions.

```{r}
# create a dataframe
plot_data <- data.frame(
  fitted = model_0$fitted.values,
  residuals = model_0$residuals
)

# build the plot

ggplot(plot_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "steelblue") +   # the points     
  geom_hline(yintercept = 0, # horizontal reference line
             linetype = "dashed", 
             color = "black") +
  geom_smooth(method = "loess", # The trend line (similar to the red line in base plot)
              formula = 'y ~ x',
              color = "orange", 
              se = FALSE) +
  theme_minimal() +
  labs(title = "Residuals vs Fitted after Log Transformation",
       x = "Fitted Values",
       y = "Residuals")
```

There is evidence of heteroscedasticity as the points are clustered around the left and fan out towards the right. This tells that there is a violation in the constant variance assumption of linearity.

Let's try a transformed model using the logarithm

```{r}
# create a dataframe
plot_data_1 <- data.frame(
  fitted = model_1$fitted.values,
  residuals = model_1$residuals
)

# build the plot

ggplot(plot_data_1, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5, color = "steelblue") +   # the points     
  geom_hline(yintercept = 0, # horizontal reference line
             linetype = "dashed", 
             color = "black") +
  geom_smooth(method = "loess", # The trend line (similar to the red line in base plot)
              formula = 'y ~ x',
              color = "orange", 
              se = FALSE) +
  theme_minimal() +
  labs(title = "Residuals vs Fitted",
       x = "Fitted Values",
       y = "Residuals")
```

Even after log transformation, the relationship being linear assumption is violated

Let's check for collinearity between the numerical independent variables

To find a good model that shows this relationship between the dependent and independent variables, we will use the 'forward step wise' variable selection method to select predictor variables that will be good for our model.

```{r}
set.seed(123) # set a pseudorandom number generator
index <- sample(1:nrow(pj_data), 0.3 * nrow(pj_data)) # sample 30% of the data for testing
# split the data
training_data <- pj_data[-index, ]
testing_data <- pj_data[index, ]

# perfom forward stepwise selection
model.bestsubset <- regsubsets(`Age at enrollment` ~.,data = pj_data, nvmax = 36) # nvmax = 36 because we have 36 independent variables
summary(model.bestsubset)
```
