---
title: "Predicting_Students_Dropout_and_Academic_Success"
author: "James Kibiro, Josh Balderas, Neelkumar Patel"
format: html
editor: visual
---

## Predicting Students Dropout and Academic Success

Research Question: What are the biggest contributors to student academic dropout and failure in higher education?

```{r}
# set up the libraries
library(tidyverse)
library(readxl) 
library(here) # sets path to your folder directory
library(leaps) # for variable selection methods
library(ggplot2)
library(caret)
library(nnet) # for multinomial logistic regression
library(pROC) # for ROC curve
library(car) # for stress-testing regression models
```

```{r}
# load the data
pj_data <- read_xlsx(here("data", "project_data.xlsx")) # the data file is in a folder named 'data'
head(pj_data)
str(pj_data)
summary(pj_data)
```

```{r}
# remove any missing values
pj_data <- na.omit(pj_data) 
```

```{r}
# Factorize the categorical variables
# 'Target' is the only categorical variable
col <- c("Target")
pj_data [, col] <- lapply(pj_data[, col], factor)
str(pj_data) # confirm categorization of the column
levels(pj_data$Target)
```

For our research question, "What are the biggest contributors to student academic dropout and failure in higher education?", we will take 'Target' as the dependent (Y) variable that we can predict using the other variables as the independent variables.

```{r}
# Let's take a look at the distribution of our variable of interest
pj_data |>
  ggplot(aes(x =Target, fill = Target))+
           geom_bar()+
  theme_minimal()+
  labs( title = "Distribution of Student Outcomes",
        x = "Outcome",
        y = "Count")
```

Let's develop a logistic regression model to predict student academic outcome.

Since Target is a categorical variable (3 classes: Dropout, Graduate, Enrolled), we cannot use linear regression as it would require a continuous numeric outcome. Instead, we will use a logistic regression (mulitnomial).

Since the outcome (Y) is 'Target' what we want to predict; Let's split the data 70% training and 30% testing

```{r}
set.seed(123)  # setting a seed for reproducibility of the random split
index <- createDataPartition(pj_data$`Target`, p = 0.3, list = FALSE)
training_data <- pj_data[-index, ]
testing_data <- pj_data[index, ]
```

Check summary statistics of the outcome "Target" in training and testing data

```{r}
summary(training_data$`Target`)
summary(testing_data$`Target`)
```

Confirm sizes of the testing and training data

```{r}
cat("Training rows:", nrow(training_data), "\n")
cat("Testing rows:", nrow(testing_data), "\n")
```

Confirm class balance is preserved in both sets

```{r}
prop.table(table(training_data$Target))
prop.table(table(testing_data$Target))
```

Before building the model, let's check whether any of our independent variables are highly correlated with each other. If two variables are almost identical, it will confuse the model and inflate the standard errors.

We will use the Variance Inflation Factor(VIF), whereby if vif = 1, we have no correlation at atll, VIF is 1- 5, we moderate correlation (fine), and VIF 5 - 10 we have a higher correlation.

```{r}
# We first fit a temporary linear model just to extract VIF values - this is a common workaround since vif() requires a linear model object
temp_model <- lm(as.numeric(Target) ~ ., data = training_data)
vif_values <- vif(temp_model) # get the vif values from the model
vif_values[vif_values > 5] # show variables that have vif values greater than 5 (potential collinearity issue)
```

We have 10 independent variables that have vif values above 5 where 6 of the variables have vif values above 10. For the variables with a vif above 10 we will remove one of them say 'Curricular units 1st sem (enrolled)'.

We can now develop a logistic regression model to predict 'Target' using all other independent variables. Since Target has 3 categories (Dropout, Enrolled, Graduate), we will use multinomial logistic regression i.e., the model will pick one class as the reference/ baseline - by default this will be 'Dropout' as it's the category that comes first alphabetically.

```{r}
model_0 <- multinom(Target~., data = training_data, maxit = 500)
# maxit = 500 gives the optimizer more iterations to converge
summary(model_0)
```

Let's check the model fit. We will use AIC and Pseudo R-squared since a logistic regression doesn't have R-squared. Rule of thumb, a lower AIC is better and a McFadden's Pseudo R squared of between 0.2 - 0.4 is considered a good fit for logistic regression.

```{r}
# model fit: 
AIC(model_0) # lower AIC is better

# McFadden's Pseudo R-squared
# Formula: 1 - (log-likelihood of full model / log-likelihood of null model)
null_model <- multinom(Target ~ 1, data = training_data, maxit = 500)
mcfadden_r2 <- 1 - (logLik(model_0) / logLik(null_model))
cat("McFadden's Pseudo R²:", round(as.numeric(mcfadden_r2), 4), "\n") # print the pseudo R squared value
```

Our AIC is 3542.901, we will use this as the baseline once we start comparing with other models. Nevertheless, our McFadden R-squared is 0.46, on a scale of 0 to 1, this is a strong model considering an excellent model has values between 0.2 and 0.4.

Let's evaluate the model using a confusion matrix

```{r}
# make predictions on the test data
predicted_class <- predict(model_0, newdata = testing_data) 

# confusion matrix
conf_matrix <- confusionMatrix(predicted_class, testing_data$Target)
conf_matrix
```
