---
title: "Predicting_Students_Dropout_and_Academic_Success"
author: "James Kibiro, Josh Balderas, Neelkumar Patel"
format: html
editor: visual
---

## Predicting Students Dropout and Academic Success

### Research Question: What are the biggest contributors to student academic dropout and failure in higher education?

#### Libraries

```{r}
# set up the libraries
library(tidyverse)
library(readxl) 
library(here) # sets path to your folder directory
library(leaps) # for variable selection methods
library(ggplot2)
library(caret)
library(nnet) # for multinomial logistic regression
library(pROC) # for ROC curve
library(car) # for stress-testing regression models
library(rpart) # for regression trees
library(rpart.plot) # for regression tree plot
library(vip) # for feature importance
library(randomForest)
library(janitor) # to clean messy column names
library(knitr) # to convert from qmd to r script
```

### Data exploration

```{r}
# load the data
pj_data <- read_xlsx(here("data", "project_data.xlsx")) 
# the data file is in a folder named 'data'
head(pj_data)
str(pj_data)
summary(pj_data)
```

```{r}
# clean messy column names
pj_data <- clean_names(pj_data) 

# remove any missing values
pj_data <- na.omit(pj_data) 
```

```{r}
colnames(pj_data)
```

```{r}
# Factorize the categorical variables
# 'target' is the only categorical variable
col <- c("target")
pj_data [, col] <- lapply(pj_data[, col], factor)
str(pj_data) # confirm categorization of the column
levels(pj_data$target)
table(pj_data$target)
```

For our research question, "What are the biggest contributors to student academic dropout and failure in higher education?", we will take 'target' as the dependent (Y) variable that we can predict using the other variables as the independent variables.

```{r}
# Let's take a look at the distribution of our variable of interest
pj_data |>
  ggplot(aes(x =target, fill = target))+
           geom_bar()+
  theme_minimal()+
  labs( title = "Distribution of Student Outcomes",
        x = "Outcome",
        y = "Count")
```

### Logistic Regression Model

Let's develop a logistic regression model to predict student academic outcome.

Since target is a categorical variable (3 classes: Dropout, Graduate, Enrolled), we cannot use linear regression as it would require a continuous numeric outcome. Instead, we will use a logistic regression (mulitnomial).

Also, for a multinomial logistic regression we need to identify the reference level in the dependent variable. Based on our research question, ' What contributes to dropout?', we will set 'Graduate' as the reference level in the 'target' variable.

```{r}
# set reference level
pj_data$target <- relevel(pj_data$target, ref = "Graduate" )

# confirm
levels(pj_data$target)
```

Since the outcome (Y) is 'target' what we want to predict; Let's split the data 70% training and 30% testing

#### Model training and testing

```{r}
set.seed(123)  # setting a seed for reproducibility of the random split
index <- createDataPartition(pj_data$`target`, p = 0.3, list = FALSE)
training_data <- pj_data[-index, ]
testing_data <- pj_data[index, ]
```

Check summary statistics of the outcome "target" in training and testing data

```{r}
summary(training_data$`target`)
summary(testing_data$`target`)
```

Confirm sizes of the testing and training data

```{r}
cat("Training rows:", nrow(training_data), "\n")
cat("Testing rows:", nrow(testing_data), "\n")
```

Confirm class balance is preserved in both sets

```{r}
prop.table(table(training_data$target))
prop.table(table(testing_data$target))
```

Before building the model, let's check whether any of our independent variables are highly correlated with each other. If two variables are almost identical, it will confuse the model and inflate the standard errors.

We will use the Variance Inflation Factor(VIF), whereby,

- if VIF = 1, we have no correlation at all,

- if VIF = 1- 5, we have moderate correlation (fine), and

- if VIF = 5 - 10 we have a higher correlation.

```{r}
# We first fit a temporary linear model just to extract VIF values - this 
# is a common workaround since vif() requires a linear model object
temp_model <- lm(as.numeric(target) ~ ., data = training_data)
vif_values <- vif(temp_model) # get the vif values from the model
vif_values[vif_values > 5] # show variables that have vif values greater 
# than 5 (potential collinearity issue)
```

- We have 10 independent variables that have VIF values above 5 where 6 of the variables have VIF values above 10.

- For the variables with a VIF value above 5 we will compare them pairwise using a correlation matrix and pick the one that is more meaningful from each correlated pair.

```{r}
# Compute a correlation matrix for all numeric predictors
# (excluding Target)
numeric_vars <- training_data |> select(-target)
cor_matrix <- cor(numeric_vars)

# Visualize it — easier to spot pairs of correlated variables
library(corrplot)
corrplot(cor_matrix, method = "color", tl.cex = 0.6)
```

Based on the correlation plot, we will remove some of the variables and keep the more meaningful one from each correlated pair.

```{r}
# Parental background — keep mothers_qualification & fathers_occupation
# Previous qualification — keep previous_qualification_grade
# Curricular units — keep 2nd sem approved & grade only
# Economic indicators — keep unemployment_rate

vars_to_remove <- c(
  "fathers_qualification",
  "mothers_occupation",
  "previous_qualification",
  "curricular_units_1st_sem_credited",
  "curricular_units_1st_sem_enrolled",
  "curricular_units_1st_sem_evaluations",
  "curricular_units_1st_sem_approved",
  "curricular_units_1st_sem_grade",
  "curricular_units_1st_sem_without_evaluations",
  "curricular_units_2nd_sem_credited",
  "curricular_units_2nd_sem_enrolled",
  "curricular_units_2nd_sem_evaluations",
  "curricular_units_2nd_sem_without_evaluations",
  "inflation_rate",
  "gdp"
)

training_data <- training_data |> select(-all_of(vars_to_remove))
testing_data  <- testing_data  |> select(-all_of(vars_to_remove))
```

Rerun VIF to check and confirm

```{r}
temp_model_2 <- lm(as.numeric(target) ~ ., data = training_data)
vif(temp_model_2)
# All VIF values should now be below 5
```

Every single VIF value is below 5. We now have 22 variables in our training data in which 21 are independent variables.

We can now develop a logistic regression model to predict 'target' using all other independent variables.

```{r}
model_0 <- multinom(target~., data = training_data, maxit = 500)
# maxit = 500 gives the optimizer more iterations to converge
summary(model_0)
```

Let's check the model fit. We will use AIC and Pseudo R-squared since a logistic regression doesn't have R-squared. Rule of thumb, a lower AIC is better and a McFadden's Pseudo R squared of between 0.2 - 0.4 is considered a good fit for logistic regression.

```{r}
# model fit: 
AIC(model_0) # lower AIC is better

# McFadden's Pseudo R-squared
# Formula: 1 - (log-likelihood of full model / log-likelihood of null model)
null_model <- multinom(target ~ 1, data = training_data, maxit = 500)
mcfadden_r2 <- 1 - (logLik(model_0) / logLik(null_model))
cat("McFadden's Pseudo R²:", round(as.numeric(mcfadden_r2), 4), "\n") # print
# the pseudo R squared value
```

Our AIC is 4101.955, we will use this as the baseline once we start comparing with other models. In addition, our McFadden R-squared is 0.3641, on a scale of 0 to 1, this is a good fit considering an excellent model has values between greater than 0.4. This means 36% of the variation in student outcomes can be explained by the model.

To get our biggest contributors to student dropout (**THIS ANSWERS OUR RESEARCH QUESTION**), we will exponentiate the coefficients from the model to get the Odds ratios - also known as Relative Risk Ratios(RRR).

- If RRR (Odds ratio) = 1.0 -\> no effect

- if RRR (Odds ratio) \> 1.0 -\> that variable increases the probability of being in that class vs the reference class (Graduate)

- if RRR (Odds ratio) \< 1.0 -\> that variable increases the probability of being in that class vs the reference class (Graduate)

For example: if the odd ratio for "Curricular units 1st sem (approved)" in the Graduate vs. Dropout row is 1.8, it means a student who passes one more unit in the first semester is 1.8× more likely to dropout than to graduate.

```{r}
# Odds ratios a.k.a Relative Risk Ratios: exp(coefficients)
odds_ratios <- exp(coef(model_0))
odds_ratios
```

#### Variable importance: Which predictors matter most?

We will use the variable importance plot to make the odds ratio easier to interpret.

```{r}
# Create a clean odds ratio dataframe for plotting
or_df <- data.frame(
  Variable = colnames(coef(model_0)),
  OR_Dropout = as.numeric(exp(coef(model_0))["Dropout", ])
) |>
  filter(Variable != "(Intercept)") |>  # remove intercept
  mutate(
    Direction = ifelse(OR_Dropout > 1, "Risk Factor", "Protective Factor")
  ) |>
  arrange(desc(OR_Dropout))

# Plot
ggplot(or_df, aes(x = reorder(Variable, OR_Dropout), 
                   y = OR_Dropout, 
                   fill = Direction)) +
  geom_col() +
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "black") +
  coord_flip() +
  scale_fill_manual(values = c("Risk Factor" = "tomato", 
                                "Protective Factor" = "steelblue")) +
  theme_minimal() +
  labs(title = "Variable Importance Plot",
    subtitle = "Odds Ratios: Dropout vs. Graduate",
       x = "Variable",
       y = "Odds Ratio",
       fill = "")
```

Everything to the right of the dashed line (0R\>1) = increases dropout risk compared to graduating. These are all the variables shown in red. Our strongest risk factor to dropout is debt. Being a debtor (financial obligation) makes a student 2.6x more likely to dropout than graduate.

Everything to the left of the dashed line (OR\<1) = decreases dropout risk compared to graduating. These are all the variables shown in blue. Having tuition fees up to date (financial stability) is the strongest positive factor. Students with fees paid are 94% less likely to drop out.

Clearly financial factors dominate both sides.

- A student who is a debtor and does not have fees paid is by far the highest dropout risk profile.

- Financial support such as education bursaries, scholarships, etc are the single most powerful intervention point.

- Academic performance in the 2nd semester (units approved) is the second most important domain.

- Demographics like gender, age, special needs play a role but are less dominant than financial and academic factors.

#### Model Evaluation

Let's evaluate the model using a confusion matrix

```{r}
# make predictions on the test data
predicted_class <- predict(model_0, newdata = testing_data) 

# confusion matrix
conf_matrix <- confusionMatrix(predicted_class, testing_data$target)
conf_matrix
```

From the confusion matrix results we see that our model is able to correctly predict the outcome 74.4% of the time (accuracy) while catching each outcome positively in each class (sensitivity/ recall) as follows: 92.4% of the time in the Graduate Class, 76.4% of the time in the Dropout Class, and 20.9% of the time in the Enrolled class. This low sensitivity in the 'Dropout' Class is likely due to the ambiguous nature of this class since they have not yet graduated or dropped out so their characteristics overlap with both other classes.

### Classification tree model: (using method= "class")

```{r}
# classification tree model
model1 <- rpart(formula = target ~ .-target, 
                data= training_data, 
                method= "class")
# plot the model
rpart.plot(model1)
head(training_data)
```

- Most important variable: Curricular_units_2nd_sem_approved, it appears at the very first split.

- Financials factor is still critical, 'tuition_fees_up_to_date' appears on both branches. Even good students drop out without it.

- Age matters only for struggling students. Age_at_enrollment only appears deeper in the tree for high risk students.

- The tree confirms our logistic regression model since both models agree on the same top predictors i.e., curriucular_units_2nd_sem_approved (strongest positive factor), tuition_fees_up_to_date (strongest protective factor), age_at_enrollment(mild risk factor)

#### Model Evaluation: Model 1

```{r}
# make predictions on the test data
predicted_class_1 <- predict(model1, newdata = testing_data) 

testing_data$predicted_class_1 <- predict(model1, newdata = testing_data, 
                                          type = "class")
head(testing_data)

# confusion matrix
conf_matrix_1 <- confusionMatrix(testing_data$predicted_class_1, 
                                 testing_data$target)
conf_matrix_1

```

From the results, we see our model is 75.5% accurate in predicted which students will graduate! In addition, our kappa is 0.59 which is a moderate to substantial measure that our accuracy isn't just a fluke of having more 'Graduate' data points.

Our Model's sensitivity is 89% for 'Graduate Category' which means 89% of the time, it positively identifies students who will graduate, and 76% of the time it will identify students who will drop out or at high risk of dropping out which is a solid detection rate. Nevertheless, the model struggles to capture students who fall in the 'Enrolled class' (37.2%) - this is most likely because the 'Enrolled Class' is a middle ground category, so its features overlap with both other categories.

Comparing the logistic regression model to the classification model, the classification model wins! The decision tree (classification model) has a higher accuracy (75.5% vs 74.4%), higher enrolled balanced accuracy (64.4% vs 57.9%).

#### Variable importance Plot

```{r}
vip(model1)+
  geom_col(fill="steelblue")+
  labs( title = "Variable Importance Plot",
        subtitle = "Classification Model")+
  theme_minimal()
```

### Random Forests (RF)

Recall: Arguments of randomForest() function

- mtry = nr of variables to pick each time we build a regression tree

- ntree = how many trees we want

- maxnodes = how many nodes we want to keep

- importance enables the algorithm to calculate variable importance

#### RF model 1

```{r}
set.seed(123) # for reproducibility
rf_model1 <- randomForest(target ~ ., 
                          data=training_data, 
                          mtry = 5,
                          importance=TRUE)

# The default mtry for classification is the square root of the 
# number of predictors. We have 21 predictors, therefore 5 would 
# be a good value for mtry.

rf_model1 
```

Our RF model has an overall OOB error rate of 24.72% meaning it is able to correctly classify 75.28% students.

The high error rate for the **Enrolled** class (73.3%) stands out. While the model is great at identifying students who graduate, it is struggling significantly to distinguish between students who have Dropped out and those who have Graduated.

Let's check how the Target class is balanced as any inbalance in the 'Target' categories could be causing this high error rate!

```{r}
table(training_data$target)
```

The table confirms a significant **class imbalance**: we have nearly three times as many Graduates as Enrolled students. The model is biased toward the "Graduate" class because it has the most observations (1,546). The "Enrolled" class is the smallest, leading the model to misclassify 256 Enrolled students as Graduates!

Let's try a **Balanced Random Forest** by sampling an equal number of rows from each class when building the trees.

#### Balanced RF model 1

```{r}
# Let's set the sample size based on our smallest class (Enrolled = 555)
# This forces the model to treat all classes with equal weight
rf_model1_balanced <- randomForest(
  target ~ ., 
  data = training_data, 
  mtry = 5, 
  importance = TRUE,
  sampsize = c("Graduate" = 555, "Dropout" = 555, "Enrolled" = 555)
)

print(rf_model1_balanced)
```

This is a much better balance! By downsampling, we cut the **Enrolled** error rate from **73% down to 51%**.

While the overall out-of-bag (OOB) error rose slightly (from 24.72% to 25.95%), the model is now significantly more "honest" about the Enrolled class. It's no longer just defaulting to the majority labels.

Looking at the confusion matrix:

- **Enrolled vs. Graduate:** 182 Graduates were mistaken for Enrolled, and 181 Enrolled were mistaken for Graduates. These two classes are still bleeding into each other, likely because their academic data (grades/credits) looks very similar.

- **The Trade-off:** The **Graduate** error doubled (from \~7% to 15%). This is the standard "price" of balancing; the model is no longer biased toward the majority, so it makes more mistakes there to get the minority class right.

#### Model Evaluation

```{r}
# predict using the balanced random forest model
predicted_class_3 <- predict(rf_model1_balanced, newdata = testing_data)

# confusion matrix
brf_cm <- confusionMatrix(predicted_class_3, testing_data$target)
brf_cm
```

Our Enrolled sensitivity is now nearly 54% (53.56%) which means we are correctly identifying nearly 54% of the Enrolled Students. Our accuracy is 77% (76.8) which is higher than our unbalanced random forest model (75.28) and also higher than our logistic regression (74.4%) and decision tree models (75.5%). Lastly, our Dropout class prediction is high 84.4%, meaning it is incredibly good at not falsely accusing a student of dropping out if they aren't actually doing so.

Using the importance() function, we can view the importance of each importance() variable

#### Variable Importance

```{r}
varImpPlot(rf_model1_balanced)
importance(rf_model1_balanced)

# variable importance plot
vip(rf_model1_balanced)+
  geom_col(fill = "steelblue")+
  labs( title = "Variable Importance Plot",
        subtitle = "Balanced Random Forest Model")+
  theme_minimal()
```

1.  Looking at the variable importance plot and data, 'curricular_units_2nd_sem_approved' is the most important variable in predicting outcomes for our model. It has a massive Mean Decrease Gini of 199.96 and Mean Decrease Accuracy of 79.08, meaning it is the most critical variable in predicting whether a student graduates or drops out.

    In addition, the top predictors are almost all related to the 2nd semester:

    - curricular_units_2nd_sem_approved

    - curricular_units_2nd_sem_grade

      If a student is failing to get credits approved in the second semester, the model is very confident they will drop out.

2.  Tuition is the most important non-academic variable. It has a high Dropout specific accuracy score (43.3), suggesting that being behind on fees is a huge red flag for immediate dropout.

3.  The 'Enrolled' column in the importance table has much lower values than other classes columns. This explains why our 'Enrolled' error is still high! The model has great features of identifying student success (Graduate) or failure (Dropout) but the features aren't very good at identifying enrolled students i.e., the 'in-between'.

4.  Variables such as 'international', 'educational_special_needs', and 'application_order' have near zero importance. They are adding noise to our random forest.

Recap: When running the variable importance under our logistic regression model, the variable 'international' came up as the most important variable for predicting the outcome. However, we have seen with the random forest model that this variable 'international' only adds noise to the model! This is because of two things: non linear relationships and the importance scale.

- Nonlinear vs Linear: Logistic regression looks for a straight line relationship. Therefore, if being an international student has a consistent, small impact across the board, then the regression model might flag it as highly significant. However, the Random Forest, looks for 'splits'. If 'curricular_units_2nd_sem_approved' is a much more powerful way to divide the data into groups, the Random Forest will prioritize it and ignore 'weaker' signals like 'international'.

- Importance Scale: In our variable importance plot in the logistic regression, our scale of importance only goes up to 4 whereas in the Random Forest, the Gini importance for academic units went up to 199. This suggests that while 'international' is statistically significant in a regression, its actual predictive power is tiny compared to grades and credits!

### Conclusion

Looking at the models above: Multinomial logistic regression, Classification tree, and Random Forest, there is a general theme of variables with 'grades' having more predictive power in determining whether a student will graduate.

Looking at the model results, the Balanced Random Forest has the highest overall accuracy (76.8%) of all the models, but it also the worst at noting that the 'Enrolled' group (Enrolled sensitivity: 53.6%) is higher than in the other models but it's the lowest in 'Graduate sensitivity': 86.3%. The Balanced Random Forest is not the best in predicting the majority class (Graduate sensitivity': 86.3%) but it does best in not ignoring the minority sensitivity (Enrolled sensitivity: 53.6%).

Looking at the balanced accuracy tells the truth: The balanced accuracy for the Enrolled class in the logistic model is 81.6%, which is lower than the Classification tree model (83.5%) and much lower than that of our Balanced Random Forest (84.3%). Therefore, the Balanced Random Forest is much better as it has a higher accuracy for the minority class across the different models.

The balanced Random Forest is our most fair and useful model as it provides the best 'safety net' for the Enrolled students without sacrificing too much performance on Graduates and Dropouts. The logistic regression remains the most accurate overall.

#### Model Comparison

```{r}
# Let's create a data frame with the Sensitivity scores we've gathered
comparison_data <- data.frame(
  Model = c("Multinomial Logistic", "Classification Tree", 
            "Balanced Random Forest"),
  Graduate = c(0.925, 0.890, 0.863),
  Dropout = c(0.764, 0.759, 0.752),
  Enrolled = c(0.209, 0.372, 0.536) # RF scores from the balanced OOB
)

# Reshape for plotting
plot_data <- pivot_longer(comparison_data, 
                          cols = c("Graduate", "Dropout", "Enrolled"), 
                          names_to = "Class", 
                          values_to = "Sensitivity")

# Generate a comparison plot
ggplot(plot_data, aes(x = Class, y = Sensitivity, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Model Comparison: Sensitivity by Student Outcome",
       subtitle = "The Balanced Random Forest is the only model that 
       effectively 'catches' Enrolled students",
       y = "Sensitivity (Recall)",
       x = "Student Status") +
  scale_fill_brewer(palette = "Set1") +
  geom_text(aes(label = round(Sensitivity, 2)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            size = 3)

```

### Convert to R script

```{r}
purl("project_code.qmd")
```
